The Proofs

### 2023-11-27 oopsla24/tex/05-proof.tex
Finish proofs

\begin{lemma*}[Soundness of (Positive) Anti-Unification]
    \label{lemma:au-soundness}
     Assuming $[[iP1]]$ and $[[iP2]]$ are normalized,
    if $[[Γ ⊨ iP1 ≈au iP2 ⫤ (Ξ, uQ, aus1, aus2)]]$
    then 
    \begin{enumerate*}
        \item $[[Γ ; Ξ ⊢ uQ]]$,
        \item $[[Γ ; · ⊢ aus1 : Ξ]]$ for $i \in \{1,2\}$
        are anti-unification substitutions (in particular, $[[Ξ]]$ contains only negative algorithmic variables), and
        \item $[[ [ausi] uQ = iPi ]]$ for $i \in \{1,2\}$.
    \end{enumerate*}
\end{lemma*}

@@

To prove the correctness properties of the anti-unification algorithm,
one extra observation is essential. The algorithm relies on the fact that
in the resulting tuple $[[(Ξ, uQ, aus1, aus2)]]$,
there are no two different placeholders $[[β̂⁻]]$
mapped to the same pair of types by $[[aus1]]$ and $[[aus2]]$.
This is important to guarantee that, for example, 
the anti-unifier of $[[↓↑Int → ↑Int]]$ and 
$[[↓↑Bool → ↑Bool]]$ is $[[↓â⁻ → â⁻]]$, but not
(less specific) $[[↓â⁻ → b̂⁻]]$.

@@

    \begin{lemma*}[Characterization of Normalized Supertypes]
        \label{lemma:char-supertypes}
        For a normalized positive type $[[iP]]$ well-fordmed in $[[Γ]]$,
        the set of normalized $[[Γ]]$-formed supertypes of $[[iP]]$ is the following:
        \begin{itemize}
            \item if $[[iP]]$ is a variable $[[pb]]$, its only normalized supertype is $[[pb]]$ itself;
            \item if $[[iP]]$ is an existential type $[[ ∃nbs.iP' ]]$ then 
                its $[[Γ]]$-formed supertypes are the $([[G, nbs]])$-formed supertypes of $[[iP']]$ not using $[[nbs]]$;
            \item if $[[iP]]$ is a downshift type $[[↓iM]]$, 
                its supertypes have form $[[∃nas.↓iM']]$ such that there exists
                a $[[Γ]]$-formed instantiation of $[[nas]]$ in $[[↓iM']]$
                that makes $[[↓iM']]$ equal to $[[↓iM]]$, \ie $[[ [iNs/nas] ↓iM' = ↓iM ]]$.
        \end{itemize}
    \end{lemma*}

    Similarly to the Least Upper Bound algorithm, the Upgrade finds the least type among upper bounds
    (this time the set of considered upper bounds consists of supertypes well-formed in a \emph{smaller} context).
    This way, we also use the supertype characterization to prove the following property of the Upgrade algorithm.

@@

    \begin{lemma*}[Upgrade Completeness]
        Assuming $[[iP]]$ is well-formed in $[[Γ = Δ, pnas]]$,
        for any $[[iQ']]$ such that $[[iQ']]$ is a $[[Δ]]$-formed upper bound of $[[iP]]$, \ie
        \begin{enumerate*}
            \item $[[Δ ⊢ iQ']]$ and
            \item $[[Γ ⊢ iQ' ≥ iP]]$,
        \end{enumerate*}
        the result of the upgrade algorithm $[[iQ]]$ exists
        ($[[upgrade Γ ⊢ iP to Δ = iQ]]$) and satisfies $[[Δ ⊢ iQ' ≥ iQ]]$.
    \end{lemma*}

@@

    The soundness of positive subtyping states that
    the output constraint $[[SC]]$ provides a sufficient set of restrictions.
    In other words, any substitution satisfying $[[SC]]$, 
    ensures the desired declarative subtyping: $[[Γ ⊢ [uσ]uP ≥ iQ]]$.
    Notice that the soundness requires that only the left-hand side input 
    type ($[[uP]]$) is algorithmic. This is one of the invariants of the algorithm
    that significantly simplifies the unification and constraint resolution.

\begin{lemma*}[Soundness of the Positive Subtyping]
    If $[[Γ ⊢ Θ]]$, $[[Γ ⊢ iQ]]$, $[[Γ ; dom(Θ) ⊢  uP]]$, and 
    $[[Γ ; Θ ⊨ uP ≥ iQ ⫤ SC]]$,
    then $[[Θ ⊢ SC : uv uP]]$ and
    for any  $[[uσ]]$ such that $[[ Θ ⊢ uσ : SC ]]$,
    we have $[[ Γ ⊢ [uσ]uP ≥ iQ ]]$.
\end{lemma*}

@@

\begin{lemma*}[Completeness of the Negative Subtyping]
    Suppose that $[[Γ ⊢ Θ]]$, $[[Γ ⊢ iM]]$, $[[Γ ; dom(Θ) ⊢ uN]]$,
    and $[[uN]]$ does not contain negative unification variables ($[[α̂⁻]] \notin [[uv uN]]$).
    Then for any $[[Θ ⊢ uσ : uv(uN)]]$ such that $[[Γ ⊢ [uσ]uN ≤ iM]]$,
    the subtyping algorithm succeeds: $[[Γ ; Θ ⊨ uN ≤ iM ⫤ SC]]$.
\end{lemma*}

\subsection{Singularity}

@@

Finally, we discuss the proofs of the theorems stated at the beginning of this section.

\paragraph{Determinicity}
    One of the properties that our proof relies on is the determinicity of the 
    typing algorithm: the output (the type, the set of restrictions, \etc) 
    is uniquely determined by the input (the term, the context, \etc).
    Determinicity is not hard to demonstrate by structural induction: 
    in every algorithmic inference system, only one inference rule can be applied 
    for a given input.  However, we need to prove it for \emph{every} 
    subroutine of the algorithm. Ultimately, it requires 
    the determinicity of such procedures as \emph{generation of fresh variables}, 
    which is easy to ensure, but must be taken into account in the implementation.

\begin{lemma*}[Determinicity of the Typing Algorithm]
    Suppose that $[[Γ ⊢ Φ]]$ and $[[Γ ⊢ Θ]]$. Then 
    \begin{itemize}
        \item [$+$] If $[[Γ; Φ ⊨ v : iP]]$ and $[[Γ; Φ ⊨ v : iP']]$ then $[[iP = iP']]$.
        \item [$-$] If $[[Γ; Φ ⊨ c : iN]]$ and $[[Γ; Φ ⊨ c : iN']]$ then $[[iN = iN']]$.
        \item If $[[Γ; Φ; Θ ⊨ uN ● args ⇒> uM ⫤ Θ'; SC]]$ and 
            $[[Γ; Φ; Θ ⊨ uN ● args ⇒> uM' ⫤ Θ'; SC']]$ then 
            $[[uM = uM']]$, $[[Θ]] = [[Θ']]$, and $[[SC]] =[[SC']]$.  
    \end{itemize}
\end{lemma*}

\paragraph{Mutual Induction of Soundness and Completeness}

The soundness and completeness of type inference cannot be proven separately:
the inductive proof of one requires another, and vice versa. 
Soundness proof requires completeness in the case of 
\ruleref{\ottdruleATAppLetLabel},
The soundness proof can be considered as a transformation from the algorithmic 
inference tree to a declarative one. In the case of \ruleref{\ottdruleATAppLetLabel},
the declarative rule requires that the (declaratively) inferred type of the application
is unique.  To show that, we take an arbitrary application typing tree and apply the typing
\emph{completeness}. Then the soundness of singularity

>>>>

\begin{lemma*}[Soundness of (Positive) Anti-Unification]
    \label{lemma:au-soundness}
     Assuming $[[iP1]]$ and $[[iP2]]$ are normalized,
    if $[[Γ ⊨ iP1 ≈au iP2 ⫤ (Ξ, uQ, aus1, aus2)]]$
    then 
    \begin{enumerate*}
        \item $[[Γ ; Ξ ⊢ uQ]]$,
        \item $[[Γ ; · ⊢ ausi : Ξ]]$ for $i \in \{1,2\}$
        are anti-unification substitutions (in particular, $[[Ξ]]$ contains only negative algorithmic variables), and
        \item $[[ [ausi] uQ = iPi ]]$ for $i \in \{1,2\}$.
    \end{enumerate*}
\end{lemma*}

@@

To prove the correctness properties of the anti-unification algorithm,
one extra observation is essential. The algorithm relies on the fact that
in the resulting tuple $[[(Ξ, uQ, aus1, aus2)]]$,
there are no two different placeholders $[[β̂⁻]]$
mapped to the same pair of types by $[[aus1]]$ and $[[aus2]]$.
This is important to guarantee that, for example, 
the anti-unifier of $[[↓↑Int → ↑Int]]$ and 
$[[↓↑Bool → ↑Bool]]$ is $[[↓â⁻ → â⁻]]$, but not
(less specific) $[[↓â⁻ → b̂⁻]]$.
To preserve this property, we arrange the algorithm in such a way 
that the name of the placeholder is determined by the types it is mapped to.
The following lemma specifies this observation.

@@

    \begin{lemma*}[Characterization of Normalized Supertypes]
        \label{lemma:char-supertypes}
        For a normalized positive type $[[iP]]$ well-formed in $[[Γ]]$,
        the set of normalized $[[Γ]]$-formed supertypes of $[[iP]]$ is the following:
        \begin{itemize}
            \item if $[[iP]]$ is a variable $[[pb]]$, its only normalized supertype is $[[pb]]$ itself;
            \item if $[[iP]]$ is an existential type $[[ ∃nbs.iP' ]]$ then 
                its $[[Γ]]$-formed supertypes are the $([[G, nbs]])$-formed supertypes of $[[iP']]$ not using $[[nbs]]$;
            \item if $[[iP]]$ is a downshift type $[[↓iM]]$, 
                its supertypes have form $[[∃nas.↓iM']]$ such that there exists
                a $[[Γ]]$-formed instantiation of $[[nas]]$ in $[[↓iM']]$
                that makes $[[↓iM']]$ equal to $[[↓iM]]$, \ie $[[ [iNs/nas] ↓iM' = ↓iM ]]$.
        \end{itemize}
    \end{lemma*}

    Similarly to the Least Upper Bound algorithm, the Upgrade finds the least type among upper bounds
    (this time the set of considered upper bounds consists of supertypes well-formed in a \emph{smaller} context).
    This way, we also use the supertype characterization to prove the following properties of the Upgrade algorithm.

@@

    \begin{lemma*}[Upgrade Completeness]
        Assuming $[[iP]]$ is well-formed in $[[Γ = Δ, pnas]]$,
        for any $[[iQ']]$ such that $[[iQ']]$ is a $[[Δ]]$-formed upper bound of $[[iP]]$, \ie
        \begin{enumerate*}
            \item $[[Δ ⊢ iQ']]$ and
            \item $[[Γ ⊢ iQ' ≥ iP]]$,
        \end{enumerate*}
        there exists$[[iQ]]$ such that
        ($[[upgrade Γ ⊢ iP to Δ = iQ]]$) and $[[Δ ⊢ iQ' ≥ iQ]]$.
    \end{lemma*}

@@

    The soundness of positive subtyping states that
    the output constraint $[[SC]]$ provides a sufficient set of restrictions.
    In other words, any substitution satisfying $[[SC]]$, 
    ensures the desired declarative subtyping: $[[Γ ⊢ [uσ]uP ≥ iQ]]$.
    Notice that the soundness requires that only the left-hand side input 
    type ($[[uP]]$) can have algorithmic variables. This is one of the invariants of the algorithm
    that significantly simplifies the unification and constraint resolution.

\begin{lemma*}[Soundness of the Positive Subtyping]
    If $[[Γ ⊢ Θ]]$ and $[[Γ ⊢ iQ]]$ and $[[Γ ; dom(Θ) ⊢  uP]]$ and 
    $[[Γ ; Θ ⊨ uP ≥ iQ ⫤ SC]]$,
    then $[[Θ ⊢ SC : uv uP]]$ and
    for any  $[[uσ]]$ such that $[[ Θ ⊢ uσ : SC ]]$,
    we have $[[ Γ ⊢ [uσ]uP ≥ iQ ]]$.
\end{lemma*}

@@

\begin{lemma*}[Completeness of the Negative Subtyping]
    Suppose that $[[Γ ⊢ Θ]]$ and $[[Γ ⊢ iM]]$ and $[[Γ ; dom(Θ) ⊢ uN]]$
    and $[[uN]]$ does not contain negative unification variables ($[[α̂⁻]] \notin [[uv uN]]$).
    Then for any $[[Θ ⊢ uσ : uv(uN)]]$ such that $[[Γ ⊢ [uσ]uN ≤ iM]]$,
    the subtyping algorithm succeeds: $[[Γ ; Θ ⊨ uN ≤ iM ⫤ SC]]$.
\end{lemma*}

\subsection{Singularity}
\label{sec:singularity-proof}

@@

Finally, we discuss the proofs of the soundness and completeness of type
inference algorithm that we stated at the beginning of this section.
There are three subtleties that we will cover that are important 
for the proof to go through: the determinacy of the algorithm,
the mutual dependence of the soundness and completeness proofs, and
the non-trivial inductive metric that we use.

\paragraph{Determinacy}
    One of the properties that our proof relies on is the determinacy of the 
    typing algorithm: the output (the inferred type)
    is uniquely determined by the input (the term and the contexts).
    Determinacy is not hard to demonstrate by structural induction: 
    in every algorithmic inference system, only one inference rule can be applied 
    for a given input.  However, we need to prove it for \emph{every} 
    subroutine of the algorithm. Ultimately, it requires 
    the determinacy of such procedures as \emph{generation of fresh variables}, 
    which is easy to ensure, but must be taken into account in the implementation.

\begin{lemma*}[Determinacy of the Typing Algorithm]
    Suppose that $[[Γ ⊢ Φ]]$ and $[[Γ ⊢ Θ]]$. Then 
    \begin{itemize}
        \item [$+$] If $[[Γ; Φ ⊨ v : iP]]$ and $[[Γ; Φ ⊨ v : iP']]$ then $[[iP = iP']]$.
        \item [$-$] If $[[Γ; Φ ⊨ c : iN]]$ and $[[Γ; Φ ⊨ c : iN']]$ then $[[iN = iN']]$.
        \item If $[[Γ; Φ; Θ ⊨ uN ● args ⇒> uM ⫤ Θ'; SC]]$ and 
            $[[Γ; Φ; Θ ⊨ uN ● args ⇒> uM' ⫤ Θ'; SC']]$ then 
            $[[uM = uM']]$, $[[Θ]] = [[Θ']]$, and $[[SC]] =[[SC']]$.  
    \end{itemize}
\end{lemma*}

\paragraph{Mutuality of the Soundness and Completeness Proofs}

Typically in our inductive proofs, the soundness is proven before completeness,
as the completeness requires certain properties of the premise subtrees.
However, in the case of the typing algorithm, the soundness and completeness proofs
cannot be separated: the inductive proof of one requires another, and vice versa.

The soundness proof can be viewed as a mapping from an algorithmic tree to a
declarative one. We show that each algorithmic inference rule can
be transformed into the corresponding declarative rule, as long as the premises
are transformed accordingly, and apply the induction principle.

The soundness requires completeness in the case of
\ruleref{\ottdruleATAppLetLabel}. To prove the soundness, 
in other words, to transform this rule into its
declarative counterpart \ruleref{\ottdruleDTAppLetLabel}, one needs
to prove \emph{uniqueness} of the inferred application type $[[↑iQ]]$. 
To show the uniqueness, we need to conclude that any other \emph{declarative} tree
(which is given by the induction hypothesis) infers an equivalent type. 
The only way to do so is by applying the soundness of singularity (\cref{sec:singularity-proo}), however, first, the declarative tree must be converted to the algorithmic one (by completeness!).
This way, the soundness and completeness proofs are mutually dependent.

\paragraph{Inductive Metric}
    The soundness and completeness lemmas are proven by \emph{mutual} induction. 
    Since the declarative and the algorithmic systems do not depend on each other,
    we must introduce a uniform \emph{metric} on which the induction is conducted.
    We define the metric gradually, starting from the
    auxiliary function---the size of a judgment $\size{J}$.

    The size of \emph{declarative} and \emph{algorithmic} judgments is defined as
    a pair: the first component is the syntactic size of the terms used in the judgment. 
    The second component depends on the kind of judgment. For regular
    type inference judgments (such as $[[Γ ; Φ ⊢ v : iP]]$
    or $[[Γ ; Φ ⊨ c : iN]]$), it is always zero. For application inference judgments
    $([[Γ ; Φ ⊢ iN ● args ⇒> iM]]$ or $[[Γ ; Φ ; Θ ⊨ uN ● args ⇒> uM ⫤ Θ'; SC]]$), 
    it is equal to the number of prenex quantifiers of the head type $[[iN]]$.
    We need this adjustment to ensure the monotonicity of the metric, since the rules 
    \ruleref{\ottdruleATForallAppLabel} and \ruleref{\ottdruleDTForallAppLabel}
    only reduce the quantifiers in the head type but not change the list of arguments.

\begin{definition*}[Judgement Size]
    \label{def:decl-typing-size}
    For a declarative or an algorithmic typing judgment $J$, we define a metric $\size{J}$ as a pair of integers in the following way:

    \begin{multicols}{2}
    \begin{itemize}
        \item [$+$] $\size{[[Γ ; Φ ⊢ v : iP]]} = (\size{[[v]]}, 0)$;
        \item [$-$] $\size{[[Γ ; Φ ⊢ c : iN]]} = (\size{[[c]]}, 0)$;
        \item [$\bullet$] $\size{[[Γ ; Φ ⊢ iN ● args ⇒> iM]]}=$\\ 
            $(\size{[[args]]}, \npq{[[iN]]})$;
    \end{itemize}
    \columnbreak
    \begin{itemize}[leftmargin=*]
        \item [$+$] $\size{[[Γ ; Φ ⊨ v : iP]]} = (\size{[[v]]}, 0)$;
        \item [$-$] $\size{[[Γ ; Φ ⊨ c : iN]]} = (\size{[[c]]}, 0)$;
        \item [$\bullet$] $\size{[[Γ ; Φ ; Θ ⊨ uN ● args ⇒> uM ⫤ Θ'; SC]]} =$\\ 
             $(\size{[[args]]}, \npq{[[uN]]})$.
    \end{itemize}
    \end{multicols}

    Here $\size{[[v]]}$ and $\size{[[c]]}$ is the size of the 
    syntax tree of the term,
    and $\size{[[args]]}$ is the sum of sizes of the terms in $[[args]]$;
    and $\npq{[[iN]]}$ and $\npq{[[iP]]}$ represent the number of 
    prenex quantifiers, \ie
        \begin{itemize} \centering
            \item [$+$] $\npq{[[∃nas.iP]]} = |[[nas]]|$, if $[[iP ≠ ∃nbs.iP']]$,
            \item [$-$] $\npq{[[∀pas.iN]]} = |[[pas]]|$, if $[[iN ≠ ∀pbs.iN']]$.
        \end{itemize}
\end{definition*}

Notice that for \emph{algorithmic} inference system, $\size{J}$ decreases in
all the inductive steps, \ie for each inference rule, the size of the
premise judgments is strictly less than the size of the conclusion.
However, the \emph{declarative} inference system has rules
\ruleref{\ottdruleDTNEquivLabel} and \ruleref{\ottdruleDTPEquivLabel}, that
`step to' an equivalent type, and thus, technically, might keep the judgment
unchanged altogether.

To deal with this issue, we introduce the metric on the entire 
\emph{inference trees} rather than on judgments, and plug into this metric the
parameter that certainly decreases in rules \ruleref{\ottdruleDTNEquivLabel} and \ruleref{\ottdruleDTPEquivLabel}---the number of these rules in the inference tree. We
denote this number as $\eqNodes{T}$.  Then the final metric is defined as a pair in the following way.

\begin{definition*}[Inference Tree Metric]
    For a tree $T$, inferring a declarative or an algorithmic judgement $J$, we define $\metric{T}$ as follows:
        \[
        \metric{T} = \begin{cases}
        (\size{J}, \eqNodes{T}) & \text{if } J \text{ represents a declarative judgement}, \\
        (\size{J}, 0) & \text{if } J \text{ represents an algorithmic judgement}.
        \end{cases}
        \]
    Here $\eqNodes{T}$ is the number of nodes in $T$ labeled with \ruleref{\ottdruleDTPEquivLabel} or \ruleref{\ottdruleDTNEquivLabel}.
\end{definition*}

This metric is suitable for mutual induction on the soundness and completeness
of the typing algorithm.  First, it is monotonous \wrt the inference rules,
and this way, we can always apply the induction hypothesis to premises 
of each rule. Second, the induction hypothesis is powerful enough,
so we can use the completeness of the algorithm in the soundness proof,
where required. For instance, to prove the soundness
of typing in case of $[[Γ; Φ ⊨ let x = v(args); c' : iN]]$,
we can assume, that  \emph{completeness}
holds for a term of shape $[[Γ; Φ ⊢ iM ● args ⇒> iK ]]$, since
$\size{args} < \size{let x = v(args); c'}$.
This is exactly what allows us to 
deal with the case of \ruleref{\ottdruleATAppLetLabel},
because then we can conclude that 
the inferred type (of a declarative judgment $[[Γ; Φ ⊢ iM ● args ⇒> ↑[uσ]uQ
]]$ constructed by the induction hypothesis) is unique.

### 2023-11-27 proofs/tex/decl-typing-lemmas.tex
Finish proofs

\begin{definition}[Number of Equivalence Nodes]
    For a tree $T$ inferring
    a declarative typing judgment,
    let us a function $\eqNodes{T}$
    as the number of nodes in $T$ labeled with \ruleref{\ottdruleDTPEquivLabel} or 
    \ruleref{\ottdruleDTNEquivLabel}.
\end{definition}

>>>>

\begin{definition}[Number of Equivalence Nodes]
    For a tree $T$ inferring
    a declarative typing judgment,
    let us define a function $\eqNodes{T}$
    as the number of nodes in $T$ labeled with \ruleref{\ottdruleDTPEquivLabel} or 
    \ruleref{\ottdruleDTNEquivLabel}.
\end{definition}
